---
title: "CodeforLecture22"
output: html_document
date: "2026-04-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Two Sided Hypothesis Testing

```{r}
library(ps270data)
library(tidyverse)

data(social, package = "qss")
```
data

```{r}
social <- as.tibble(social)
social
```
Permutating labels
```{r}
social <- social |>
  filter(messages %in% c("Neighbors", "Control"))

social |>
  mutate(messages_permute = sample(messages)) |>
  select(primary2006, messages, messages_permute)
```

Let's check how many are in each arm
```{r}
table(social$messages)
```

## Two-sample permutation tests with infer
calculating the diff in proportion

```{r}
library(infer)
social <- social |>
  mutate(turnout = if_else(primary2006 == 1, "Voted", "Didn't Vote"))

```

```{r}
est_ate <- social |>
  specify(turnout ~ messages, success = "Voted") |>
  calculate(stat = "diff in props", order = c("Neighbors", "Control"))
est_ate
```

Setting the hypothesis
```{r}
social |>
  specify(turnout ~ messages, success = "Voted") |>
  hypothesize(null = "independence")
```
Generating the permulations

```{r}
social |>
  specify(turnout ~ messages, success = "Voted") |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute")
```
bootstrap is randomly choosing samples
permutations is assigning random treatment in a new column

calculate the difference in proportions in each sample using the infer package instead of group_by

```{r}
null_dist <- social |>
  specify(turnout ~ messages, success = "Voted") |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in props", order = c("Neighbors", "Control"))
```

```{r}
null_dist |>
  visualize()
```

calculating the p-values

```{r}
ate_pval <- null_dist |>
  get_p_value(abs_stat = est_ate, direction = "both")
ate_pval
```

visualizing p-value

```{r}
null_dist |>
  visualize() +
  shade_p_value((abs_stat = est_ate, direction = "both"))
```

